Evolution of the AGN UV luminosity function from redshift 7.5

### 2018-04-30 Paper/paper.tex
Update

Several counter-arguements against a higher incidence of AGN at high
redshift have also been made. \citet{2017MNRAS.468.4691D} and
\citet{2018MNRAS.473.1416M} pointed out that an AGN population
consistent with the findings of \citet{2015AA...578A..83G} leads to a
much earlier \HeII\ reionization.  For instance, in the model of
\citet{2015ApJ...813L...8M}, He~\textsc{ii} reionization is complete
at $z=4.5$, compared to $z=3$ in the standard scenario
\citep{2012ApJ...746..125H}.  Such early \HeII\ reionization could
result in higher IGM temperatures due to the associated photoheating.
\citet{2017MNRAS.468.4691D} found that the temperature of the IGM at
mean density is twice as much in AGN-dominated reionization models as
the standard models at $z=3.5$--$5$, in conflict with constraints from
the \lya forest.  This inconsistency could be avoided by postulating a
reduced escape fraction of \HeII-ionizing photons in AGN, but it is
difficult to reconcile this with a unit escape fraction of
hydrogen-ionizing photons that is required to explain the Ly$\alpha$
opacity fluctuations.  Further arguments against AGN-dominated
reionization have been presented by \citet{2016MNRAS.459.2299F} by
analysing the incidence of metal-line absorbers at $z\sim 6$.
\citet{2016MNRAS.459.2299F} find that in their cosmological radiation
hydrodynamical simulations AGN-dominated UV background results in too
many C~\textsc{iv} absorption systems relative to Si~\textsc{iv} and
C~\textsc{ii} at $z\sim 6$.  However, these simulations assume a
$L_\nu\propto\nu^{-1.57}$ AGN SED at extreme UV
\citep{2002ApJ...579..500T}, which somewhat harder than recent
measurements ($L_\nu\propto\nu^{-1.7}$) by
\citet{2015MNRAS.449.4204L}.  \citet{2016MNRAS.459.2299F} also found
that the N(Si~\textsc{iv})/N(C~\textsc{iv}) column density ratio
measurements prefer a harder and more intense $>4$~Ry background than
the standard model \citep{2012ApJ...746..125H}.  Finally, comparing
the \citet{2015AA...578A..83G} sample to X-ray-selected quasar data at
$z=0$--$6$, \citet{2017MNRAS.465.1915R} argued that the faint end of
the AGN UV luminosity function at $z\sim 6$ is probably shallower that
that reported by \citet{2015AA...578A..83G}.
\citet{2017MNRAS.465.1915R} argue that the apparent contradiction with
the results of \citet{2015AA...578A..83G} could be explained by
contamination from the AGN host galaxies for faint AGN.

@@

\begin{table*}
  % In published version of this table, we could remove file names and add data homogenisation in Notes.
  % Add z ~ 7 qso when you add it to analysis
  % See data.tex for a version of this table that includes file names
  % Should we mark Giallongo, Mortlock, and Banados entries to say that these are sometimes excluded from analysis?
  % Remove notes?
  % Do we really need a sample ID column?
  % Say in caption or text that sample 13 is restricted to $z<4.7$
  % Also that sample 17 overlaps with 8 for $M_{1450}<-26.73$
  % Also that in sample 7 we rescaled selection probabilities 
  \caption{Quasar data sets}
  \label{tab:samples}
  \begin{tabular}{crcllrS}
    \hline
    & ID & $z$ & Survey & Reference & Number & {Area} \\
    & & & & & of quasars & {(deg$^2$)} \\
    \hline
    1. & 13 & 0.1--2.2 & SDSS DR7 & \citet{2006AJ....131.2766R} & 48664 & 6248.0 \\
    2. & 15 & 0.4--2.6 & 2SLAQ SGP & \citet{2009MNRAS.392...19C} & 3663 & 64.2 \\
    3. & 15  & 0.4--2.6 & --- NGP & \citet{2009MNRAS.392...19C} & 8153 & 127.7 \\
    4. &  1 & 2.2--3.5 & BOSS DR9 & \citet{2013ApJ...773...14R} & 23301 & 2236.0 \\
    5. & 13 & 3.7--4.7 & SDSS DR7 & \citet{2006AJ....131.2766R} & 1785 & 6248.0 \\
    6. & 17 & 4.7--5.4 & SDSS+WISE & \citet{2016ApJ...829...33Y} & 99 & 14555.0 \\
    7. &  8 & 4.7--5.1 & SDSS DR7 & \citet{2013ApJ...768..105M} & 148 & 6248.0 \\
    8. &  8 & 4.7--5.1 & --- Stripe 82 & \citet{2013ApJ...768..105M} & 52 & 235.0 \\
    9. &  8 & 5.1--5.5 & --- DR7 Extended & \citet{2013ApJ...768..105M} & 28 & 6248.0 \\
    10. & 8 & 5.1--5.4 & ---  Stripe 82 Extended & \citet{2013ApJ...768..105M} & 10 & 235.0 \\
    11. & 6 & 3.7--4.8 & NDWFS & \citet{2011ApJ...728L..26G} & 12 & 1.71 \\
    12. & 6 & 3.8--5.1 & DLS & \citet{2011ApJ...728L..26G} & 12 & 2.05 \\
    13. & 7 & 4.1--6.3 & CANDELS GOODS-S & \citet{2015AA...578A..83G} & 19 & 0.047 \\
    14. & 18 & 5.8--6.4 & --- Main & \citet{2016ApJ...833..222J} & 24 & 11240.0 \\
    15. & 18 & 5.9--6.1 & --- Overlap & \citet{2016ApJ...833..222J} & 10 & 4223.0 \\
    16. & 18 & 5.7--6.1 & --- Stripe 82 & \citet{2016ApJ...833..222J} & 13 & 277.0 \\
    17. & 10 & 6.0 & CFHQS Deep & \citet{2010AJ....139..906W} & 1 & 4.47 \\
    18. & 10 & 5.9--6.4 & --- Very Wide & \citet{2010AJ....139..906W} & 16 & 494.0 \\
    19. & 11 & 6.0--6.2 & Subaru High-$z$ Quasar & \citet{2015ApJ...798...28K} & 2 & 6.5 \\
    20. & 19 & 6.5--7.4 & UKIDSS & \citet{2011Natur.474..616M} & 1 & 3370.0 \\
    21. & 19 & 6.5--7.4 & UKIDSS & \citet{2007MNRAS.376L..76V} & 1 & 3370.0 \\
    22. & 20 & 6.5--7.4 & ALLWISE+UKIDSS+DECaLS & \citet{2018Natur.553..473B} & 1 & 2500.0 \\
    \hline
  \end{tabular}
\end{table*}

@@

We started with compiling the samples of recent UV-optical quasar
surveys. The restriction to UV-optical surveys was mainly driven by
our science goal to characterise the UV luminosity function of Type~I
quasars, but also by the still uncertain conversion from X-ray flux to
UV flux, which adds unnecessary systematics for large
samples. \textbf{References!} The individual surveys and their main
characteristics are listed in
Table~\ref{tab:samples}. Figure~\ref{fig:qsos} presents a redshift
histogram of the contributing surveys.

We included surveys based on a set of simple criteria:
\begin{enumerate}
\item Spectroscopic redshifts for the majority of targets.
\item Accurate rest-frame UV-optical CCD photometry.
\item Statistical power (sample size, coverage in $z$ and/or $M_{1450}$).
\item A characterised selection function.
\end{enumerate}
As a prerequisite for a joint analysis of the QLF we obtained the
survey selection functions in electronic form, either from the
publication or from the authors. As a reference for future surveys we
publish them here in modified and homogenised form
(Section~\ref{sect:datahom}). \textbf{Maybe get the authors' consent.}

Due to their selection criteria and their statistical power specific
surveys contribute to distinct redshift ranges. At $z<2.2$ we
considered quasars from the SDSS DR7 quasar catalogue
\citep{2010AJ....139.2360S} and the 2SLAQ catalogue
\citep{2009MNRAS.392...19C}. We restricted the SDSS DR7 sample to XXX
$z<2.2$ quasars selected with the final SDSS quasar selection
algorithm \citep{2002AJ....123.2945R, 2006AJ....131.2766R} from a
survey area of 6248\,deg$^2$ \citep{2012ApJ...746..169S}. We adopted
the SDSS targeting photometry corrected for Galactic extinction
\citep{2010AJ....139.2360S}. To limit systematic uncertainties in the
correction for host galaxy light \citep{2009MNRAS.392...19C} we
restricted the 2SLAQ sample to XXX $g<21.85$ $0.4<z<2.2$ quasars from
its spectroscopic survey footprint near the North Galactic Pole (NGP,
XX quasars in $127.7$\,deg$^2$) and the South Galactic Pole (SGP, XX
quasars in $64.2$\,deg$^2$). The small sample overlap between SDSS and
2SLAQ (112 quasars) has negligible impact on the QLF evaluation.

@@

At $3.7<z<4.1$ we used a combination of SDSS DR7 \citep[XX uniformly
  selected quasars from][]{2010AJ....139.2360S} and the NDWFS$+$DLS
survey \citep{2010ApJ...710.1498G,2011ApJ...728L..26G}. The lower cut
$z>3.7$ limits the impact of systematic uncertainties in the
\citet{2006AJ....131.2766R} SDSS selection function
\citep{2009ApJ...705L.113P, 2011ApJ...728...23W}. We did not consider
the results from surveys for faint $z\sim 4$ quasars in the COSMOS
field \citep{2011ApJ...728L..25I, 2012ApJ...755..169M} due to
systematic errors in their selection functions\footnote{Both studies
  simulated quasar colours with a mean IGM attenuation curve
  \citep{1995ApJ...441...18M} that cannot account for stochastic Lyman
  continuum absorption, and therefore underpredicts the variance in
  quasar colours \citep{1999ApJ...518..103B, 2008MNRAS.387.1681I,
    2011ApJ...728...23W}. Modelling the colour variance in these
  surveys is essential, as most of the \citet{2011ApJ...728L..25I}
  quasars are near the edge of their colour selection region (see
  their Figure~1), and \citet{2012ApJ...755..169M} require modest
  attenuation of the $U$ band flux relative to the mid-infrared
  flux.}. Furthermore, 30 per cent of the \citet{2012ApJ...755..169M}
COSMOS sample have visually estimated photometric redshifts, and the
spectroscopic subsample reveals that 40 per cent of the visually
estimated redshifts are biased low
($z_\mathrm{spec}>z_\mathrm{est}+0.3$, see their Figure~9). These
unaccounted systematic redshift errors at least partly explain the
discrepancy in the $z\sim 4$ QLF between \citet{2011ApJ...728L..26G}
and \citet{2012ApJ...755..169M}, which justifies our preference for
the former sample.

Between $z=4.1$ and $z=4.7$ our combined sample has XX uniformly
selected quasars from the SDSS DR7 quasar catalogue
\citep{2010AJ....139.2360S} and XX quasars from
\citet{2011ApJ...728L..26G}. Although we do not include the highly
debated \citet{2015AA...578A..83G} sample in our main analysis due to
its rough selection function and lacking spectroscopy for 17 of the 22
quasar candidates, we use it in Section~XX to constrain the faint end
($M_{1450}>-23$) of the QLF at $z>4.1$. We restricted the
\citet{2015AA...578A..83G} sample to the 19/22 sources considered in
their QLF.

At $4.7\le z<5.1$ we combined several recent surveys, accounting for
sample overlap and updated selection functions. At the bright end of
the QLF we used the 71 $z<5.1$ quasars from the SDSS+WISE survey
\citep{2016ApJ...829...33Y} together with its selection function.
%Although \citet{yang} are sensitive to $z>5.1$ quasars, we limit
%their sample to $z<5.1$ to limit differential evolution of the QLF
%across the redshift range.
To avoid double-counting quasars contained both in
\citet{2016ApJ...829...33Y} and the SDSS DR7 sample from
\citet{2013ApJ...768..105M}, we used the latter only at
$M_{1450}>-26.72$ (in our adopted cosmology), yielding 94 additional
$4.7\le z<5.1$ quasars selected in 6248\,deg$^2$. We used the updated
$z\sim 5$ SDSS DR7 selection function \citep{2013ApJ...768..105M}
instead of the one from \citet{2006AJ....131.2766R}.
%\footnote{We note that the major difference in the selection
%functions must be due to different input parameters, since the single
%color selection criterion in \citet{mcgreer} selects the vast
%majority of the SDSS DR7 quasars.}.
To these two bright-end samples we added the faint-end sample from the
\citet{2013ApJ...768..105M} SDSS Stripe~82 survey (52 uniformly
selected $4.7\le z<5.1$ quasars in 235\,deg$^2$) and two quasars from
\citet{2011ApJ...728L..26G}, adopting the respective selection
functions. We did not consider the limit on the $z\sim 5$ QLF by
\citet{2012ApJ...756..160I} due to systematic errors in their
selection function\footnote{\citet{2012ApJ...756..160I} underestimated
  the dispersion in rest-frame UV quasar colours with respect to SDSS
  at all redshifts (their Figure~4). Contrary to their claim,
  photometric errors have a small effect on the colour distribution of
  SDSS quasars given the statistical errors of $<0.03$\,mag in $gri$
  for 90 per cent of the SDSS DR7 bright quasar sample ($i<19.1$) and
  a relative calibration error of $\sim 1$ per cent
  \citep{2008ApJ...674.1217P}.}.

The SDSS colours of $5.1\le z<5.5$ quasars are similar to those of M
and L dwarf stars, resulting in a low and uncertain selection
efficiency \citep{2013ApJ...768..105M}. WISE mid-infrared selection
performs better \citep{2016ApJ...829...33Y}, but is restricted to the
bright end of the quasar population. Our combined sample contains the
remaining 28 quasars from \citet{2016ApJ...829...33Y} and 10 quasars
from the \citet{2013ApJ...768..105M} Stripe~82 survey. As we will show
in Section~XX, the resulting QLF is consistent with those at lower and
higher redshifts, indicating that the \citet{2013ApJ...768..105M}
selection functions are quite reliable. \textbf{Check this!}

Finally, we combined the samples from all spectroscopic $z\sim 6$
quasar surveys with a determined selection function as of June
2017. \citet{2016ApJ...833..222J} recently compiled all quasars
discovered in several SDSS $z\sim 6$ surveys together with
consistently derived selection functions. Their uniform sample
consists of 24 quasars from the SDSS main survey (11240\,deg$^2$), 10
additional quasars in regions with two or more SDSS imaging scans
(so-called overlap regions, 4223\,deg$^2$), and 13 faint quasars from
SDSS Stripe~82 (277\,deg$^2$). The CFHQS \citep{2010AJ....139..906W}
provided a uniform sample of 16 quasars in the Very Wide Survey
(494\,deg$^2$) and a single quasar in the Deep Survey
($4.47$\,deg$^2$). The one quasar detected in both SDSS and CFHQS does
not lead to underestimated statistical errors in the QLF. Lastly, we
included the two objects from \citet{2015ApJ...798...28K}. One might
be a Lyman break galaxy due to its narrow Ly$\alpha$ emission line
(half width at half maximum 427\,km\,s$^{-1}$), but
\citet{2015ApJ...798...28K} note that 10/17 candidates do not have
follow-up spectroscopy, hence their QLF may be underestimated. Since
very small samples preclude a statistical correction via the selection
function, complete spectroscopic follow-up of $z\sim 6$ candidates is
essential. Although we will account for the slightly different
redshift sensitivities for the different surveys, we will quote a
nominal redshift range $5.7<z<6.5$ for the combined $z\sim 6$ sample.

@@

For a joint fit of the QLF it is necessary to homogenise the different
survey samples in absolute magnitude, and to convert their selection
functions to the same absolute magnitude system. For the analysis of
the quasar UV emissivity and to be consistent with published work at
$z>3$ we chose to convert all samples and selection functions to the
absolute AB magnitude at a rest frame wavelength $\lambda=1450$\,\AA
\begin{equation}\label{eq:absmag}
  M_{1450}\left(z\right) = m-5\log{\left(\frac{d_L\left(z\right)}{\mathrm{Mpc}}\right)}-25-K_{m,1450}\left(z\right),
\end{equation}
with the luminosity distance $d_L\left(z\right)$ to a quasar at
redshift $z$, the apparent magnitude $m$ in a filter used in the
survey, and the bandpass correction $K_{m,1450}\left(z\right)$
\citep{1956AJ.....61...97H, 1968ApJ...154...21O, 2000A&A...353..861W,
  2002astro.ph.10394H}. For the bandpass correction we used a
combination of the \citet{2015MNRAS.449.4204L} quasar SED at
$\lambda<2500$\,\AA, and the \citet{2001AJ....122..549V} quasar
composite spectrum at longer wavelengths to cover the lowest
redshifts.  The samples from SDSS and BOSS are defined in the SDSS $i$
band, while 2SLAQ is defined in the $g$ band. At $z>4.7$ we adopted
the SDSS $z$ band magnitude (in the following denoted $z_\mathrm{AB}$)
for SDSS DR7 quasars to avoid additional corrections due to the
Ly$\alpha$ forest. Figure~\ref{fig:kcorr} shows our bandpass
corrections for SDSS, BOSS and 2SLAQ as a function of redshift. We
ignored the luminosity dependence of the bandpass correction due to
the known anticorrelation of emission line equivalent width and
luminosity \citep{1977ApJ...214..679B}.  While the
\citet{2015MNRAS.449.4204L} SED is for luminous ($M_{1450}\simeq
-27.2$) quasars, UV composite spectra including fainter quasars
\citep{2002ApJ...565..773T, 2012ApJ...752..162S, 2014ApJ...794...75S}
give similar values, such that our bandpass corrections remain
applicable at $M_{1450}\la -24$. Empirical luminosity-dependent
bandpass corrections show a $\la 0.2$\,mag variation over $\sim 4$
orders of absolute magnitude depending on redshift and the filter, and
with a $\sim 0.2$\,mag intrinsic scatter due to individual SED
variations \citep{2013ApJ...773...14R, 2013ApJ...768..105M,
  2013A&A...551A..29P}.

>>>>

Several counter-arguements against a higher incidence of AGN at high
redshift have also been made. \citet{2017MNRAS.468.4691D} and
\citet{2018MNRAS.473.1416M} pointed out that an AGN population
consistent with the findings of \citet{2015AA...578A..83G} leads to a
much earlier \HeII\ reionization.  For instance, in the model of
\citet{2015ApJ...813L...8M}, He~\textsc{ii} reionization is complete
at $z=4.5$, compared to $z=3$ in the standard scenario
\citep{2012ApJ...746..125H}.  Such early \HeII\ reionization could
result in higher IGM temperatures due to the associated photoheating.
\citet{2017MNRAS.468.4691D} found that the temperature of the IGM at
mean density is twice as much in AGN-dominated reionization models as
the standard models at $z=3.5$--$5$, in conflict with constraints from
the \lya forest.  This inconsistency could be avoided by postulating a
reduced escape fraction of \HeII-ionizing photons in AGN, but it is
difficult to reconcile this with a unit escape fraction of
hydrogen-ionizing photons that is required to explain the Ly$\alpha$
opacity fluctuations.  Further arguments against AGN-dominated
reionization have been presented by \citet{2016MNRAS.459.2299F} by
analysing the incidence of metal-line absorbers at $z\sim 6$.
\citet{2016MNRAS.459.2299F} find that in their cosmological radiation
hydrodynamical simulations AGN-dominated UV background results in too
many C~\textsc{iv} absorption systems relative to Si~\textsc{iv} and
C~\textsc{ii} at $z\sim 6$.  However, these simulations assume a
$L_\nu\propto\nu^{-1.57}$ AGN SED at extreme UV
\citep{2002ApJ...565..773T}, which somewhat harder than recent
measurements ($L_\nu\propto\nu^{-1.7}$) by
\citet{2015MNRAS.449.4204L}.  \citet{2016MNRAS.459.2299F} also found
that the N(Si~\textsc{iv})/N(C~\textsc{iv}) column density ratio
measurements prefer a harder and more intense $>4$~Ry background than
the standard model \citep{2012ApJ...746..125H}.  Finally, comparing
the \citet{2015AA...578A..83G} sample to X-ray-selected quasar data at
$z=0$--$6$, \citet{2017MNRAS.465.1915R} argued that the faint end of
the AGN UV luminosity function at $z\sim 6$ is probably shallower that
that reported by \citet{2015AA...578A..83G}.
\citet{2017MNRAS.465.1915R} argue that the apparent contradiction with
the results of \citet{2015AA...578A..83G} could be explained by
contamination from the AGN host galaxies for faint AGN.

@@

\begin{table*}
  \caption{Data sets analysed in this work.  We restrict the SDSS DR7
    data set \citep{2010AJ....139.2360S} to $z<4.7$.  The
    \citet{2013ApJ...768..105M} data set is restricted to faint
    ($M_{1450}>-26.73$) quasars to avoid overlap with the
    \citet{2016ApJ...829...33Y} sample.  Most AGN in the
    \citet{2015AA...578A..83G} sample do not have spectroscopic
    redshifts.  We therefore only use this sample for secondary
    analysis in Section~\ref{sec:global} and Appendix~\ref{sec:conv}.
    Similarly, the quasars reported by \citet{2011Natur.474..616M},
    \citet{2015ApJ...801L..11V}, and \citet{2018Natur.553..473B} have
    only approximate completeness estimates.  We only use these
    quasars in our secondary analysis in Section~\ref{sec:global}.}
  \label{tab:samples}
  \begin{tabular}{ccllrS}
    \hline
    & $z$ & Survey & Reference & Number & {Area} \\
    & & & & of quasars & {(deg$^2$)} \\
    \hline
    1. & 0.1--2.2 & SDSS DR7 & \citet{2010AJ....139.2360S} & 48664 & 6248.0 \\
    2. & 0.4--2.6 & 2SLAQ SGP & \citet{2009MNRAS.392...19C} & 3663 & 64.2 \\
    3. &  0.4--2.6 & --- NGP & \citet{2009MNRAS.392...19C} & 8153 & 127.7 \\
    4. & 2.2--3.5 & BOSS DR9 & \citet{2013ApJ...773...14R} & 23301 & 2236.0 \\
    5. & 3.7--4.7 & SDSS DR7 & \citet{2010AJ....139.2360S} & 1785 & 6248.0 \\
    6. & 4.7--5.4 & SDSS+WISE & \citet{2016ApJ...829...33Y} & 99 & 14555.0 \\
    7. & 4.7--5.5 & SDSS DR7 & \citet{2013ApJ...768..105M} & 103 & 6248.0 \\
    8. & 4.7--5.4 & --- Stripe 82 & \citet{2013ApJ...768..105M} & 59 & 235.0 \\
    9. &3.7--4.8 & NDWFS & \citet{2011ApJ...728L..26G} & 12 & 1.71 \\
    10. &3.8--5.1 & DLS & \citet{2011ApJ...728L..26G} & 12 & 2.05 \\
    11. &4.1--6.3 & CANDELS GOODS-S & \citet{2015AA...578A..83G} & 19 & 0.047 \\
    12. & 5.8--6.4 & SDSS Main & \citet{2016ApJ...833..222J} & 24 & 11240.0 \\
    13. & 5.9--6.1 & --- Overlap & \citet{2016ApJ...833..222J} & 10 & 4223.0 \\
    14. & 5.7--6.1 & --- Stripe 82 & \citet{2016ApJ...833..222J} & 13 & 277.0 \\
    15. & 6.0 & CFHQS Deep & \citet{2010AJ....139..906W} & 1 & 4.47 \\
    16. & 5.9--6.4 & --- Very Wide & \citet{2010AJ....139..906W} & 16 & 494.0 \\
    17. & 6.0--6.2 & Subaru High-$z$ Quasar & \citet{2015ApJ...798...28K} & 2 & 6.5 \\
    18. & 6.5--7.4 & UKIDSS & \citet{2011Natur.474..616M} & 1 & 3370.0 \\
    19. & 6.5--7.4 & UKIDSS & \citet{2015ApJ...801L..11V} & 1 & 3370.0 \\
    20. & 6.5--7.4 & ALLWISE+UKIDSS+DECaLS & \citet{2018Natur.553..473B} & 1 & 2500.0 \\
    \hline
  \end{tabular}\\
\end{table*}

@@

We started with compiling the samples of recent rest-frame UV-optical
quasar surveys. The restriction to UV-optical surveys was mainly
driven by our science goal to characterise the UV luminosity function
of Type~1 quasars. X-ray-selected samples are less suited for this
purpose due to spectroscopic incompleteness and the $\sim 0.4$\,dex
scatter in the conversion from X-ray to UV luminosity
\citep{2010A&A...512A..34L, 2015MNRAS.453.1946G, 2016ApJ...819..154L}
that contributes significantly to the error budget in the UV
luminosity function of X-ray-selected samples unless rest-frame UV
photometry is incorporated \citep{2015AA...578A..83G}.  The individual
surveys and their main characteristics are listed in
Table~\ref{tab:samples}.  Figure~\ref{fig:qsos} presents a redshift
histogram of the contributing surveys.

We included surveys based on a set of simple criteria:
\begin{enumerate}
\item Spectroscopic redshifts for the vast majority of targets.
\item Accurate rest-frame UV-optical CCD photometry.
\item Statistical power (sample size, coverage in $z$ and/or absolute magnitude).
\item A characterised selection function.
\end{enumerate}
As a prerequisite for a joint analysis of the QLF we obtained the
survey selection functions in electronic form, either from the
publication or from the authors. As a reference for future surveys we
publish them here in modified and homogenised form
(Section~\ref{sect:datahom}).

Due to their selection criteria and their statistical power specific
surveys contribute to distinct redshift ranges. At $z<2.2$ we
considered quasars from the SDSS DR7 quasar catalogue
\citep{2010AJ....139.2360S} and the 2SLAQ survey catalogue
\citep{2009MNRAS.392...19C}. We restricted the SDSS DR7 sample to the
48,664 $0.1<z<2.2$ quasars selected with the final SDSS quasar
selection algorithm \citep{2002AJ....123.2945R, 2006AJ....131.2766R}
from a survey area of 6248\,deg$^2$ \citep{2012ApJ...746..169S}. We
adopted the SDSS targeting photometry corrected for Galactic
extinction \citep{2010AJ....139.2360S}. To limit systematic
uncertainties in the correction for host galaxy light
\citep{2009MNRAS.392...19C} we restricted the 2SLAQ sample to 11,816
$g<21.85$ $0.4<z<2.2$ quasars from its spectroscopic survey footprint
near the North Galactic Pole (NGP, 8153 quasars in $127.7$\,deg$^2$)
and the South Galactic Pole (SGP, 3663 quasars in
$64.2$\,deg$^2$). The small sample overlap between SDSS and 2SLAQ (112
quasars) has negligible impact on the QLF evaluation.

@@

At $3.7<z<4.7$ we used a combination of SDSS DR7 \citep[1785 uniformly
  selected quasars from][]{2010AJ....139.2360S} and the NDWFS$+$DLS
survey \citep{2010ApJ...710.1498G,2011ApJ...728L..26G}. The lower cut
$z>3.7$ in SDSS limits the impact of systematic uncertainties in the
\citet{2006AJ....131.2766R} selection function
\citep{2009ApJ...705L.113P, 2011ApJ...728...23W}. We did not consider
the results from surveys for faint $z\sim 4$ quasars in the COSMOS
field \citep{2011ApJ...728L..25I, 2012ApJ...755..169M} due to
systematic errors in their selection functions\footnote{Both studies
  simulated quasar colours with a mean IGM attenuation curve
  \citep{1995ApJ...441...18M} that cannot account for stochastic Lyman
  continuum absorption, and therefore underpredicts the variance in
  quasar colours \citep{1999ApJ...518..103B, 2008MNRAS.387.1681I,
    2011ApJ...728...23W}. Modelling the colour variance in these
  surveys is essential, as most of the \citet{2011ApJ...728L..25I}
  quasars are near the edge of their colour selection region (see
  their Figure~1), and \citet{2012ApJ...755..169M} require modest
  attenuation of the $U$ band flux relative to the mid-infrared
  flux.}. Furthermore, 30 per cent of the \citet{2012ApJ...755..169M}
COSMOS sample have visually estimated photometric redshifts, and the
spectroscopic subsample reveals that 40 per cent of the visually
estimated redshifts are biased low
($z_\mathrm{spec}>z_\mathrm{est}+0.3$, see their Figure~9). These
unaccounted systematic redshift errors at least partly explain the
discrepancy in the $z\sim 4$ QLF between \citet{2011ApJ...728L..26G}
and \citet{2012ApJ...755..169M}, which justifies our preference for
the former sample that is 77 per cent spectroscopically complete at
$R$ magnitudes $<23.5$ \citep{2011ApJ...728L..26G}.

At $4.7\le z<5.5$ we combined several recent surveys, accounting for
sample overlap and updated selection functions. At the bright end of
the QLF we used the 99 quasars from the SDSS+WISE survey
\citep{2016ApJ...829...33Y} that have $M_{1450}<-26.73$ in our adopted
cosmology.  For these 99 quasars selected from 14,555\,deg$^2$ we
adopted the \citet{2016ApJ...829...33Y} selection function.  The
\citet{2016ApJ...829...33Y} sample partially overlaps with the SDSS
DR7 sample from \citet{2013ApJ...768..105M}, so to avoid
double-counting quasars we used the latter sample only at $M_{1450}\ge
-26.73$, yielding 104 additional $4.7\le z<5.5$ quasars selected in
6248\,deg$^2$. We used the $z\sim 5$ SDSS DR7 selection function from
\citet{2013ApJ...768..105M} that supersedes the one from
\citet{2006AJ....131.2766R} due to improved $K$ corrections and IGM
parameterization.  To these two bright-end samples we added the
faint-end sample from the \citet{2013ApJ...768..105M} SDSS Stripe~82
survey (62 uniformly selected $4.7\le z<5.5$ quasars in 235\,deg$^2$)
and two $4.7\le z<5.5$ quasars from \citet{2011ApJ...728L..26G},
adopting the respective selection functions.  We did not consider the
limit on the $z\sim 5$ QLF by \citet{2012ApJ...756..160I} due to
systematic errors in their selection
function\footnote{\citet{2012ApJ...756..160I} underestimated the
  dispersion in rest-frame UV quasar colours with respect to SDSS at
  all redshifts (their Figure~4). Contrary to their claim, photometric
  errors have a small effect on the colour distribution of SDSS
  quasars given the statistical errors of $<0.03$\,mag in $gri$ for 90
  per cent of the SDSS DR7 bright quasar sample ($i<19.1$) and a
  relative calibration error of $\sim 1$ per cent
  \citep{2008ApJ...674.1217P}.}.

The SDSS colours of $5.1<z<5.5$ quasars are similar to those of M and
L dwarf stars, resulting in a low and uncertain selection efficiency
\citep{2013ApJ...768..105M}. WISE mid-infrared selection performs
better \citep{2016ApJ...829...33Y}, but is restricted to the bright
end of the quasar population. Unlike \citet{2013ApJ...768..105M} we
include the 9 uniformly selected $M_{1450}\ge -26.73$ SDSS DR7 quasars
and the 10 SDSS Stripe~82 quasars at $z>5.1$, adopting their low
selection efficiency.  As we will show in Section~3.2, the resulting
QLF is consistent with those at lower and higher redshifts, indicating
that the \citet{2013ApJ...768..105M} selection functions are quite
reliable.

At $z\sim 6$ we combined the samples from all spectroscopic surveys
with a determined selection function as of June 2017.
\citet{2016ApJ...833..222J} recently compiled all quasars discovered
in several SDSS $z\sim 6$ surveys together with consistently derived
selection functions. Their uniform sample consists of 24 quasars from
the SDSS main survey (11,240\,deg$^2$), 10 additional quasars in
regions with two or more SDSS imaging scans (so-called overlap
regions, 4223\,deg$^2$), and 13 faint quasars from SDSS Stripe~82
(277\,deg$^2$). The CFHQS \citep{2010AJ....139..906W} provided a
uniform sample of 16 quasars in the Very Wide Survey (494\,deg$^2$)
and a single quasar in the Deep Survey ($4.47$\,deg$^2$). The one
quasar detected in both SDSS and CFHQS does not lead to underestimated
statistical errors in the QLF. Lastly, we included the two objects
from \citet{2015ApJ...798...28K}, one of which might be a galaxy due
to its narrow Ly$\alpha$ emission line (half width at half maximum
427\,km\,s$^{-1}$). With better photometry and additional spectroscopy
recently reported by \citet{2017ApJ...847L..15O} the
\citet{2015ApJ...798...28K} sample is complete.  Although we will
account for the slightly different redshift sensitivities for the
different surveys, we will quote a nominal redshift range $5.7<z<6.5$
for the combined $z\sim 6$ sample.

At the highest redshifts $z>6.5$ we considered two quasars found by
\citet{2011Natur.474..616M} and \citet{2015ApJ...801L..11V} in UKIDSS
imaging (3370\,deg$^2$), in addition to the current highest-redshift
quasar at $z=7.54$ selected from a combination of UKIDSS, WISE and
DECaLS \citep[$\sim 2500$\,deg$^2$, ][]{2018Natur.553..473B}. Although
currently only rough estimates exist concerning their selection
functions, these quasars provide constraints on the evolution of the
integrated quasar space density from $z\sim 6$ to $z\sim 7$.  We use
them in our secondary analysis in Section~\ref{sec:global}.  Although
we do not include the highly debated \citet{2015AA...578A..83G} sample
in our main analysis due to its rough selection function and lacking
spectroscopy for 17 of the 22 quasar candidates, we use it in
Section~\ref{sec:global} to constrain the faint end ($M_{1450}>-23$)
of the QLF at $z>4.1$. We restricted the \citet{2015AA...578A..83G}
sample to the 19/22 sources considered in their QLF.

@@

For a joint fit of the QLF it is necessary to homogenise the different
survey samples in absolute magnitude, and to convert their selection
functions to the same absolute magnitude system. For the analysis of
the quasar UV emissivity and to be consistent with published work at
$z>3$ we chose to convert all samples and selection functions to the
absolute AB magnitude at a rest frame wavelength $\lambda=1450$\,\AA
\begin{equation}\label{eq:absmag}
  M_{1450}\left(z\right) = m-5\log{\left(\frac{d_L\left(z\right)}
    {\mathrm{Mpc}}\right)}-25-K_{m,1450}\left(z\right),
\end{equation}
with the luminosity distance $d_L\left(z\right)$ to a quasar at
redshift $z$, the apparent magnitude $m$ in a filter used in the
survey, and the bandpass correction $K_{m,1450}\left(z\right)$
\citep{1956AJ.....61...97H, 1968ApJ...154...21O, 2000A&A...353..861W,
  2002astro.ph.10394H}. For the bandpass correction we used a
combination of the \citet{2015MNRAS.449.4204L} quasar SED at
$\lambda<2500$\,\AA, and the \citet{2001AJ....122..549V} quasar
composite spectrum at longer wavelengths to cover the lowest
redshifts.  The samples from SDSS and BOSS are defined in the SDSS $i$
band, while 2SLAQ is defined in the $g$ band. At $z>4.7$ we adopted
the SDSS $z$ band magnitude (in the following denoted $z_\mathrm{AB}$)
for SDSS DR7 quasars to avoid additional corrections due to the
Ly$\alpha$ forest. Figure~\ref{fig:kcorr} shows our bandpass
corrections for SDSS, BOSS and 2SLAQ as a function of redshift. We
ignored the luminosity dependence of the bandpass correction due to
the known anticorrelation of emission line equivalent width and
luminosity \citep{1977ApJ...214..679B}.  While the
\citet{2015MNRAS.449.4204L} SED is for luminous ($M_{1450}\simeq
-27.2$) quasars, UV composite spectra including fainter quasars
\citep{2002ApJ...565..773T, 2012ApJ...752..162S, 2014ApJ...794...75S}
give similar values, such that our bandpass corrections remain
applicable at $M_{1450}\la -24$. Empirical luminosity-dependent
bandpass corrections show a $\la 0.2$\,mag variation over $\sim 4$
orders of absolute magnitude depending on redshift and the filter, and
with a $\sim 0.2$\,mag intrinsic scatter due to individual SED
variations \citep{2013ApJ...773...14R, 2013ApJ...768..105M,
  2013A&A...551A..29P}.
